---
title: "Social Network Analysis: Node and Graph Level Statistics Part 1"
subtitle: "EPIC - SNA, Columbia University"
author: "Zack W Almquist"
institute: "University of Minnesota"
date: "June 13th, 2018"
output: 
  beamer_presentation:
    includes:
      in_header: header.tex
      after_body: footer.tex
toc: true
#toc_depth: 2
#keep_tex: true
#https://github.com/matze/mtheme
---

```{r,include=FALSE,cache=FALSE}
require(knitr)
opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE,tidy=TRUE,comment=NA,prompt=FALSE)
options(width=70)

## Packages
library(networkMethods)
library(networkdata)
library(sna)
library(network)
library(ergm)
library(xtable)
```

# Node-Level Indices

### Node-Level Indices

Introduction to classic Social Network Metrics (Positional or Node-level indices)


### Node-level Indices

\begin{textblock*}{60mm}(-5mm,-.4\textheight)
\begin{itemize}
\item Node-level index: a real- valued function of a graph and a vertex
\begin{itemize}
\item Purely structural NLIs depend only on unlabeled graph properties
\begin{itemize}
\item I.e., $f(v,G)\rightarrow \Re$
\item Invariant to node relabeling
\end{itemize}
\item Covariate-based NLIs use both structural and covariate properties
\begin{itemize}
\item I.e., $f(v,G,X)\rightarrow \Re$
\item Not labeling invariant
\end{itemize}
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{55mm}(55mm,-.4\textheight)
\begin{itemize}
\item Primary uses:
\begin{itemize}
\item Quantify properties of individual positions
\item Describe local neighborhood
\end{itemize}
\item Several common families:
\begin{itemize}
\item Centrality
\item Ego-net structure
\item Alter covariate indices
\end{itemize}
\item Centrality is the most prominent, and our focus today/lecture
\end{itemize}
\end{textblock*}


### Centrality

\begin{itemize}
\item Returning to the core question: how do individual \underline{positions} vary?
\item One manner in which positions vary is the extent to which they are ``central" in the network
\begin{itemize}
\item Important concern of social scientists (and junior high school students)
\end{itemize}
\item Many distinct concepts
\begin{itemize}
\item No one way to be central in a network - many different kinds of centrality!
\item Different types of centrality aid/hinder different kinds of actions
\item Being highly central in one respect doesn't always mean being central in other respects (although the measures generally correlate)
\end{itemize}
\end{itemize}


### Types of Centrality Measures

\begin{itemize}
\item One attempted classification by Koschutzki et al. (2005):
\begin{itemize}
\item \underline{Reach}: Centrality based on ability of ego to reach other vertices
\begin{itemize}
\item Degree, closeness
\end{itemize}
\item \underline{Flow Mediation}: Centrality based on quantity/weight of walks passing through ego
\begin{itemize}
\item Stress, betweenness
\end{itemize}
\item \underline{Vitality}: Centrality based on effect of removing ego from the network
\begin{itemize}
\item Flow betweenness (oddly), cutpoint status
\end{itemize}
\item \underline{Feedback}: Centrality of ego defined as a recursive function of alter centralities
\begin{itemize}
\item Eigenvector centrality, Bonacich Power
\end{itemize}
\end{itemize}
\end{itemize}


### Degree

\begin{textblock*}{70mm}(-5mm,-.4\textheight)
\begin{itemize}
\item Degree: number of direct ties
\begin{itemize}
\item Overall activity or extent of involvement in relation
\item High degree positions are influential, but also may be subject to a great deal of influence from others
\end{itemize}
\item Formulas:
\begin{itemize}
\item Degree (undirected):
$$d(i,Y)=\sum_{j=1}^N Y_{ij}$$
\item Indegree: $d_i(i,Y)=\sum_{j=1}^N Y_{ji}$
\item Outdegree: $d_o(i,Y)=\sum_{j=1}^N Y_{ij}$
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(60mm,-.4\textheight)

```{r,fig.align='center',out.width='.6\\textwidth',echo=FALSE}
star<-network(cbind(rep(1,7),matrix(rep(0,times=6,each=7),nc=6,nr=7)),directed=FALSE)
plot(star,main="Undirected Star",label=1:7, edge.col=rgb(0,0,0,.3),label.cex=2)

set.seed(9902211)
net<-rgraph(5)
gplot(net,main="directed network",label=1:5,edge.col=rgb(0,0,0,.3),label.cex=2)
```

\end{textblock*}

### Review: Shortest Paths

\begin{textblock*}{100mm}(-5mm,-.4\textheight)
\begin{itemize}
\item A shortest path from $i$ to $j$ is called an $i,j$ geodesic
\begin{itemize}
\item Can have more than one (but all same length, obviously)
\item The length of an $i,j$ geodesic is called the geodesic distance from $i$ to $j$
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(-5mm,-.1\textheight)

```{r,fig.align='left',out.width='.7\\textwidth',echo=FALSE,results='asis'}
set.seed(9902211)
gplot(symmetrize(net),main="Directed Network",label=1:5,edge.col=rgb(0,0,0,.3),label.cex=2)
```
\end{textblock*}

\begin{textblock*}{50mm}(40mm,0\textheight)
\tiny{
```{r,fig.align='center',out.width='.5\\textwidth',echo=FALSE,results='asis'}
g<-sna::geodist(symmetrize(net))
print(xtable(g$gdist),floating=FALSE)
```
}
\end{textblock*}

### Betweenness

\begin{textblock*}{70mm}(-5mm,-.45\textheight)
\begin{itemize}
\item Betweenness: tendency of ego to reside on shortest paths between third parties
\begin{itemize}
\item Quantifies extent to which position serves as a bridge
\item High betweenness positions are associated with ``broker" or ``gatekeeper" roles; may be able to ``firewall" information flow
\end{itemize}
\item Formula
$$b(i,Y)=\sum_{j\neq i}\sum_{k\neq l} \frac{g^{'}(j,k,l)}{g(j,k)}$$
Where $g(j,k)$ is the number of $j,k$ geodesics, $g^{'}(j,k,i)$ is the number of $j,k$ geodesics including $i$
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(65mm,-.4\textheight)

```{r,fig.align='center',out.width='.5\\textwidth',echo=FALSE}
set.seed(9902211)
plot(star,main="Undirected Star",label=1:7, edge.col=rgb(0,0,0,.3),label.cex=2)

gplot(symmetrize(net),main="directed",label=1:5,edge.col=rgb(0,0,0,.3),label.cex=2)
```

\end{textblock*}

### Closeness

\begin{textblock*}{80mm}(-10mm,-.45\textheight)
\begin{itemize}
\item Closeness: ratio of minimum distance to other nodes to observed distance to other nodes
\begin{itemize}
\item Extent to which position has short paths to other positions
\item High closeness positions can quickly distribute information, but may have limited direct influence
\item Limitation: not useful on disconnected graphs (may need to symmetrize directed graphs, too)
\end{itemize}
\item Formula
$$c(i,Y)= \frac{N-1}{\sum_{j=1}^N D(i,j)}$$
Where $D(I,j)$ is the distance from $i$ to $j$
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(70mm,-.4\textheight)
\includegraphics[width=.9\linewidth]{figures/closeness}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}


### Classic Centrality Measures Compared

\begin{textblock*}{100mm}(10mm,-.4\textheight)
\includegraphics[width=.9\linewidth]{figures/compare}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}



### Classic Centrality Measures Compared

\begin{textblock*}{100mm}(10mm,-.4\textheight)
\includegraphics[width=.9\linewidth]{figures/krackExample}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}

### Relatedness of Centrality Indices

\begin{textblock*}{60mm}(0mm,-.4\textheight)
\begin{itemize}
\item Centrality indices are strongly correlated in practice
\item Simple example: total degree versus ``complex" NLIs
\begin{itemize}
\item Squared correlations for sample UCINET data sets
\item Some diversity, but usually accounts for majority of variance
\item Theoretical insight: if you can capture degree, you can capture many other aspects of social position
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(70mm,-.4\textheight)
\includegraphics[width=.9\linewidth]{figures/correlationCent}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}



### Relating NLIs to Vertex Covariates

\begin{textblock*}{70mm}(-10mm,-.45\textheight)
\begin{itemize}
\item Common question: are NLIs related to non-structural covariates?
\begin{itemize}
\item Centrality to power or influence
\item Constraint to advancement
\item Diversity to attainment
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(70mm,-.4\textheight)
\includegraphics[width=.9\linewidth]{figures/texasEmon}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}

### ``Linear" Permutation Tests

\begin{textblock*}{55mm}(-10mm,-.45\textheight)
\small
\begin{itemize}
\item Simple, nonparametric test of association between vectors
\begin{itemize}
\tiny
\item Sometimes called ``linear" or ``vector" permutation test (or monte carlo test)
\item Tests marginal association against exchangeability null (independence conditional on marginal distributions)
\end{itemize}
\item Null interpretation: ``musical chairs" model
\begin{itemize}
\tiny
\item If we randomly switched the positions of people in the network (leaving structure as-is), what is the chance of observing a similar degree of association?
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{75mm}(40mm,-.45\textheight)
\small
\begin{itemize}
\item Monte Carlo procedure:
\begin{itemize}
\small
\item Let $x_{obs} = (f(v_1,G),\dots,f(v_N,G))$ be the observed NLI vector, w/covariate vector $y$
\item Let $t_{obs} = s(x_{obs},y)$
\item For $i$ in $1,\dots,n$
\begin{itemize}
\small
\item Let $x^{(0)}$ be a random permutation of $x_{obs}$
\item Let $t^{(i)}=s(x^{(i),y})$
\end{itemize}
\end{itemize}
\item Estimated p-values:
\begin{itemize}
\item One-seided
\begin{itemize}
\tiny
\item $\Pr(t^{(i)} \leq t_{obs} )\approx\sum_{i} I(t^{(i)} \leq t_{obs})/n$
\item $\Pr(t^{(i)}\geq t_{obs})\approx\sum_{i} I(t^{(i)}\geq t_{obs})/n$
\end{itemize}
\item Two-sided
\begin{itemize}
\tiny
\item $\Pr(|t^{(i)}|\geq|t_{obs}|)\approx\sum_{i} I(|t^{(i)}|\geq|t_{obs}|)/n$
\end{itemize}
\end{itemize}
\end{itemize}
\end{textblock*}


### Understanding the Null Model

\begin{textblock*}{100mm}(10mm,-.4\textheight)
\includegraphics[width=.9\linewidth]{figures/nullPerm}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}



### Texas SAR EMON Example

\begin{textblock*}{65mm}(-5mm,-.4\textheight)
\begin{itemize}
\item Question: do organizations in constant communication w/many alters end up more/less prominent in the decision-making process?
\begin{itemize}
\small
\item Measure ($s$): correlation of decision rank score ($y$) with degree in confirmed ``continuous communication" network ($x_{obs}$)
\item Null: no relationship between degree and decision making
\item Alternative: decision making has linear marginal relationship w/degree
\end{itemize}
\item Results
\begin{itemize}
\item $t_{obs}= 0.86$; $\Pr(|t^{(i)}\geq|t_{obs}|) \approx 3e-5$
\item Correlation this large very unlikely under null hypothesis
\item Upper tail test similar (see figure)
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(70mm,-.4\textheight)
\includegraphics[width=.9\linewidth]{figures/permDist}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}


### NLIs as Covariates

\begin{textblock*}{60mm}(-5mm,-.4\textheight)
\begin{itemize}
\item NLIs can also be used as covariates (e.g., in regression analyses)
\begin{itemize}
\item Modeling assumption: position properties predict properties of those who hold them
\item Conditioning on NLI values, so dependence doesn't matter (if no error in $G$)
\item NLIs as dependent variables are much more problematic; we'll revisit this problem when we discuss ERGs
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{60mm}(55mm,-.4\textheight)
\begin{itemize}
\item Things to keep in mind....
\begin{itemize}
\item Make sure that your theory really posits a direct relationship w/the NLI
\item NLI distributions could be quite skewed or irregular; be sure this makes sense (e.g., via analysis of residuals)
\item Multiple NLIs may be strongly correlated; may not be able to distinguish among related measures in practice
\end{itemize}
\end{itemize}
\end{textblock*}

# Graph Level Indicies 

### Graph-Level Properties

\begin{itemize}
\item Earlier, we discussed the notion of node-level indices (mainly centrality)
\begin{itemize}
\item Dealt with position of the individual within the network
\end{itemize}
\item Today, we will focus on properties at the graph level
\begin{itemize}
\item Graph-level index: $f(v,G)\rightarrow\Re$
\item Describes aggregate features of structure as a whole
\end{itemize}
\item Provide complementary insight into social structure
\begin{itemize}
\item Node-level properties tell you who's where, but graph- level properties provide the broader context
\end{itemize}
\end{itemize}


### Review Density

\begin{itemize}
\item Density: fraction of possible edges which are present
\begin{itemize}
\item Probability that a given graph edge is in the graph
\end{itemize}
\item Formulas:
\begin{itemize}
\item Undirected: $\delta = \frac{2 \sum_{i=1}^N \sum_{j=i}^N Y_{ij}}{N(N-1)}$
\item Directed: $\delta = \frac{2 \sum_{i=1}^N \sum_{j=1}^N Y_{ij}}{N(N-1)}$
\end{itemize}
\end{itemize}

**R Code**
\tiny
```{r}
undirected<-rgraph(10,mode="graph")
directed<-rgraph(10,mode="digraph")
gden(undirected,mode="graph")
gden(directed,mode="digraph")
```

### Size, Density, and Mean Degree

\begin{itemize}
\item Important fact: size, density, and mean degree are intrinsically related
\begin{itemize}
\item Formally, $d_m = \delta(N-1)$ [I.e., mean degree = density times size-1]
\item Also, $\delta=d_m/(N-1)$ [I.e., density = mean degree over size-1]
\end{itemize}
\item Simple fact, with non-obvious implications
\begin{itemize}
\item If mean degree fixed, density falls with 1/group size
\item To maintain density, have to increase degree linearly, but actors can only support so many ties!
\item Thus, growing networks become increasingly sparse over time
\begin{itemize}
\item Durkheim, Parsons, etc: modern social order depends on/produces norms of generalized exchange, since only tiny fraction of person can be directly related
\end{itemize}
\end{itemize}
\end{itemize}

### Illustration: Mean Degree Constancy and Density Decline

\tiny
```{r,echo=TRUE,fig.align='center',message=FALSE,warning=FALSE,out.width='.6\\linewidth',cache=TRUE}
library(ggplot2)
library(gridExtra)
library(networkdata)
data(addhealth)
data<-data.frame(size=sapply(addhealth,network.size),
density=sapply(addhealth,gden))
data$meandegree<-data$density*(data$size-1)

p1<-ggplot(data, aes(size,density)) + geom_point() + geom_smooth()
p2<-ggplot(data, aes(size,meandegree)) + geom_point() + geom_smooth()
grid.arrange(p1, p2, ncol=1)
```

### Beyond Density: the Dyad Census

\begin{textblock*}{70mm}(-10mm,-.45\textheight)
\begin{itemize}
\item Dyad census: a count of the number of mutual, asymmetric and null dyads in a network
\begin{itemize}
\item Mutual: $(i,j)$ and $(j,i)$
\item Asymmetric: $(i,j)$ or $(j,i)$, but not both
\item Null: neither $(i,j)$ nor $(j,i)$
\item Traditionally written as $(M,A,N)$
\end{itemize}
\item Used as ``building block"
\begin{itemize}
\item $M+A+N$ = Number of dyads
\item $2M+A$= Number of edges
\item $(M+A/2)/(M+A+N)$= Density
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{60mm}(60mm,-.3\textheight)
\tiny
```{r,out.width='.8\\linewidth',echo=FALSE,cache=TRUE}
asym<-matrix(0,nc=4,nr=4)
asym[1,4]<-1
asym[3,2]<-1
mut<-matrix(c(0,1,1,0),nc=2)
null<-matrix(c(0,0,0,0),nc=2)
par(mfrow=c(1,3),cex.main=4)
gplot(mut,mode="circle",main="Mutual")
box()
gplot(asym,mode="circle",main="Asymmetric")
box()
gplot(null,mode="circle",main="Null")
box()
```

\end{textblock*}

### Reciprocity

\begin{textblock*}{60mm}(-5mm,-.45\textheight)
\begin{itemize}
\item Reciprocity: tendency for relations to be symmetric
\item Several notions:
\begin{itemize}
\item Dyadic: probability that any given dyad is symmetric (mutual or null)
$$\frac{M+N}{M+A+N}$$
\item Edgewise: probability that any given edge is reciprocated
$$\frac{2M}{2M+A}$$
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{60mm}(65mm,-.4\textheight)
\tiny
```{r,out.width='.4\\linewidth',echo=FALSE,results='asis',fig.align='center'}
graph<-rgraph(15)
gplot(graph)
print(xtable(dyad.census(graph)))
#print(xtable(grecip(graph,measure ="edgewise")))
#print(xtable(grecip(graph,measure ="dyadic")))
```
\end{textblock*}

### Reachability

\begin{textblock*}{60mm}(-10mm,-.45\textheight)
\begin{itemize}
\item Reachability graph
\begin{itemize}
\item Digraph, $R$, based on $G$ such that $(i,j)$ is an edge in $R$ iff there exists an $i,j$ path in $G$
\begin{itemize}
\item If $G$ is undirected or fully reciprocal, $R$ will also be fully reciprocal
\end{itemize}
\item Intuitively, an edge in $R$ connects vertices which are connected in $G$
\item Strong components of $G$ (including cycles) form cliques in $R$
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{60mm}(65mm,-.4\textheight)
\includegraphics[width=.7\linewidth]{figures/reachability}
\end{textblock*}

### Hierarchy

\begin{textblock*}{70mm}(-10mm,-.45\textheight)
\begin{itemize}
\item Hierarchy: tendency for structures to be asymmetric
\item As with reciprocity, many notions; for instance\dots
\begin{itemize}
\item Dyadic Hierarchy: 1- (Dyadic Reciprocity)
\begin{itemize}
\item Intuition: extent to which dyads are asymmetric
\end{itemize}
\item Krackhard Hierarchy: $1-M/(M+A)$ in Reachability Graph
\begin{itemize}
\item Intuition: for pairs which are in a contact, what fraction are asymmetric?
\end{itemize}
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{60mm}(65mm,-.4\textheight)

```{r,out.width='.4\\linewidth',echo=FALSE,results='hide',cache=TRUE,fig.align='center',messages=FALSE}
graph<-rgraph(15,tprob=.1)
gplot(graph)
krack<-hierarchy(graph,measure ="krackhardt")
rec<-invisible(hierarchy(graph,measure ="reciprocity"))
```
```{r,echo=FALSE,results='asis'}
cat("Reciprocity\n\n")
print(xtable(matrix(rec)),include.rownames = FALSE, include.colnames = FALSE,floating = FALSE)
cat("Krackhardt\n\n")
rec<-krack
print(xtable(matrix(rec)),include.rownames = FALSE, include.colnames = FALSE,floating = FALSE)
```
\end{textblock*}

### Centralization

\begin{textblock*}{65mm}(-5mm,-.45\textheight)
\begin{itemize}
\item Centralization: extent to which centrality is concentrated on a single vertex
\item Definition dut to Freeman (1979):
$$C(G) = \sum_{i=1}^N \left(\max_v c(v,G)-c(i,G)\right) $$
\begin{itemize}
\item Defined for any centrality measure
\item Often used with degree, betweenness, closeness, etc.
\end{itemize}
\item Most centralized structure usually star network
\begin{itemize}
\item True for most centrality measures
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{60mm}(65mm,-.4\textheight)
\tiny
```{r,out.width='.65\\linewidth',echo=FALSE,fig.align='center'}
graph<-rgraph(15,tprob=.5,mode="graph")
star<-matrix(0,nc=10,nr=10); star[1,]<-1; star[,1]<-1
par(mfrow=c(2,1),mar=c(0, 0, 0, 0),cex.main=3)
gplot(graph,main="Random Network")
gplot(star,main="Star Network")
```

```{r,echo=FALSE,results='asis',cache=TRUE}
cat("Random Network\n\n")
print(xtable(cbind(centralization(graph,g=1,degree),centralization(graph,g=1,closeness))),include.rownames = FALSE,include.colnames = FALSE,floating=FALSE)
cat("Star Network\n\n")
print(xtable(cbind(centralization(star,g=1,degree),centralization(star,g=1,closeness))),include.rownames = FALSE,include.colnames = FALSE,floating=FALSE)
```
\end{textblock*}

### Centralization Versus Hierarchy

\begin{itemize}
\item Aren't centralization and hierarchy the same thing?
\item No! Two very different ideas:
\begin{itemize}
\item Hierarchy: asymmetry in interaction
\item Centralization: inequality in centrality
\end{itemize}
\item Can have centralized mutual structures, hierarchical decentralized structures
\end{itemize}



### Centralization and Team Performance

\begin{itemize}
\item Bavelas, Leavitt and others studied work teams with four structural forms:
\end{itemize}
\includegraphics[width=.8\linewidth]{figures/bavelas}

\begin{itemize}
\item Performance generally highest in centralized groups
\begin{itemize}
\item Star, ``Y" took least time, made fewest errors, used fewest messages
\end{itemize}
\item Satisfaction generally highest in decentralized groups
\begin{itemize}
\item Circle$>$Chain$>$``Y"$>$Star (but central persons had fun!)
\end{itemize}
\item A lesson: optimal performance $\neq$ optimal satisfaction ...
\end{itemize}

# References and Places for More Information 

### References and Places for More Information {.allowframebreaks}




