---
title: "Social Network Analysis: Overview Part 1"
subtitle: "EPIC - SNA, Columbia University"
author: "Zack W Almquist"
institute: "University of Minnesota"
date: "June 11th, 2018"
output: 
  beamer_presentation:
    includes:
      in_header: header.tex
      after_body: footer.tex
toc: true
#toc_depth: 2
#keep_tex: true
#https://github.com/matze/mtheme
---

```{r,include=FALSE,cache=FALSE}
require(knitr)
opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE,tidy=TRUE,comment=NA,prompt=FALSE)
options(width=60)

## Packages
library(networkMethods)
library(networkdata)
library(sna)
library(network)
library(ergm)
library(xtable)

colorLegend<-function(net,attr){
  unique(cbind(net%v%attr,plotArgs.network(net, "vertex.col", attr)))
}

```

# Format and Overview of Course

### Format of the Course

**(Most Days)**

- 1 hour lecture
- 1 hour lab
- Break (5-10 minutes)
- 1 hour lecture
- 1 hour lab


### Overview of this Course

- Day 1 - Overview of Networks and Introduction to R for analysis
- Day 2 - Network Descriptives
- Day 3 - More Network Descriptives and Introduction to Statistical Inference for Networks
- Day 4 - Introduction to Statistical Models
- Day 5 - Advanced Statistical Models: Network Autocorrelation, TERGM and SAOM

### About these notes

Portions of material adapted from courses by:

- jimi adams, University of Colorado, Denver
- Carter Butts, UCI

\&

\tiny

- David Schaefer, Arizona State University
- Ryan Light, University of Oregon
- Ann McCranie, Indiana University
- Benjamin Lind, HSE-Moscow
- Daniel McFarland, Stanford University
- David Knoke, University of Minnesota
- Garry Robins, University of Melbourne
- Krista Gile, University of Massachusetts
- Lada Adamic, University of Michigan
- Michał Bojanowski, University of Warsaw
- Mark Handcock, UCLA
- Mark Newman, University of Michigan
- James Moody, Duke University
- Martina Morris, University of Washington
- Ryan Acton, University of Massachusetts


# What is Social Network Analysis?

### What is Social Network Analysis?

- A \red{theoretical perspective} that attempts to explain how individual and group behaviors and relations are linked to patterns in society

- A set of relational \red{methods} for systematically analyzing the ties/relations between nodes/actors

\tiny{Borgatti, SP., A Mehra, DJ Brass, \& G Labianca. 2009. "Network Analysis in the Social Sciences." Science 323:892}

### What is Social Network Analysis?

- A \red{theory} and a \red{method} for social scientific analysis that relies on:
    - \emph{Structural} intuition
    - Systematic collection of \emph{relational} data   
    - Use of graphic \emph{visualizations}
    - Mathematical or computational \emph{models}

\tiny{Freeman, Linton C. 2004. The Development of Social Network Analysis. Empirical Press.}

### What is Social Network Analysis?

\begin{center}
\includegraphics[width=1\linewidth]{figures/moreno}
\end{center}

\tiny{Martina Morris. Statistical Models for Social Networks. University of Washington.}

### What is Social Network Analysis?

\includegraphics[width=1\linewidth]{figures/googleNgram.png}

### What is Social Network Analysis?

\begin{textblock*}{60mm}(-3mm,-.4\textheight)
\begin{itemize}
\item Relational (network) data concerns connections among entities, rather than attributes of entities
\begin{itemize}
\item \red{Entities} can be persons, organizations, concepts, etc.
\item \red{Relations} can be interaction, proximity, membership, etc.
\end{itemize}
\end{itemize}
\end{textblock*}

\begin{textblock*}{50mm}(60mm,-.4\textheight)
```{r,echo=FALSE,message=FALSE,warnings=FALSE,cache=TRUE,results='hide'}
data(emon)
i<-4
w<-as.sociomatrix(emon[[i]],"Frequency")
w<-(1/max(w))*w
pdf("figures/emonMt.pdf")
set.seed(99)
par(bg=NA,mar=c(0, 0, 0, 0) + 0.1)
plot(emon[[i]],
#main=names(emon)[i],
#edge.lwd="Frequency",
edge.col=rgb(0,0,0,.75),
edge.lwd=w,
displayisolates=FALSE, 
vertex.col=rainbow(8),
vertex.cex=2,
usecurve =TRUE,
displaylabels=FALSE)
#legend("left",legend=(emon[[i]]%v%"vertex.names")[degree(emon[[i]])!=0],pch=19,col=rainbow(8),bty="n",cex=1.5)
dev.off()
pdf("figures/emonMt_legend.pdf")
par(bg=NA,mar=c(0, 0, 0, 0) + 0.1)
plot.new()
legend("center",legend=(emon[[i]]%v%"vertex.names")[degree(emon[[i]])!=0],pch=19,col=rainbow(8),bty="n",cex=1.5)
dev.off()
```
\tiny{Mt. Si Search and Rescue Emergent Multi-organizational Network}
\includegraphics[width=1\linewidth]{figures/emonMt.pdf}

\end{textblock*}

\begin{textblock*}{50mm}(60mm,0.025\textheight)
\includegraphics[width=1\linewidth]{figures/emonMt_legend.pdf}
\end{textblock*}

\begin{textblock*}{100mm}(-3mm,.425\textheight)
\tiny{Drabek, T.E.; Tamminga, H.L.; Kilijanek, T.S.; and Adams, C.R. (1981). Data from Managing Multiorganizational Emergency Responses: Emergent Search and Rescue Networks in Natural Disaster and Remote Area Settings. Program on Technology, Environment, and Man Monograph 33. Institute for Behavioral Science, University of Colorado.}
\end{textblock*}

# Preliminaries: Why Networks?

### Some Types of Networks

\begin{itemize}
\item Conceptual: shared or antithetic properties
\begin{itemize}
\item E.g., similarity/difference in individual attributes, correlation among variables, inclusion/exclusion, surface matchings on proteins
\end{itemize}
\item Co-categorical: shared membership
\begin{itemize}
\item E.g., organizational co-membership, event co-participation, co-occurrence of words within texts
\end{itemize}
\item Nominational: resulting from the behavior of respondent (\red{ego})
\begin{itemize}
\item E.g., attributions of friendship/enmity, kinship (fictive or otherwise), causal narratives, reported sexual contact networks, reported needle sharing
\end{itemize}
\item Behaviorial
\begin{itemize}
\item E.g., face-to-face communication, radio communication, physical contact
\end{itemize}
\item Online Social networks
\begin{itemize}
\item E.g., Facebook, Twitter, Linkedin
\end{itemize}
\end{itemize}

### Examples!

Now let's look at some examples\dots

### Correlates of War Project: Militarized Interstate Disputes

```{r, echo=FALSE,message=FALSE,out.width='.55\\textwidth',fig.align='center'}
data(mids_1993)
ideg <- degree(mids_1993, cmode="indegree")              # Indegree for MIDs
odeg <- degree(mids_1993, cmode="outdegree")             # Outdegree for MIDs
gplot(mids_1993, vertex.cex=(ideg+odeg)^0.5/2, vertex.sides=50, 
    boxed.labels=FALSE,label.cex=0.4, 
    vertex.col=rgb(odeg/max(odeg),0,ideg/max(ideg)), 
    label=network.vertex.names(mids_1993))
title("1993 militarized interstate disputes (MIDs)")#,sub="Correlates of War Project")
```

### Correlates of War Project: Contiguity Network

```{r, echo=FALSE,message=FALSE,out.width='.5\\textwidth',fig.align='center'}
data(contig_1993)
bet <- betweenness(contig_1993, gmode="graph")       # Geographic betweenness
gplot(contig_1993, vertex.cex=sqrt(bet)/25, gmode="graph")   # Use w/gplot
title("Contiguity Network, 1993")#,sub="Correlates of War Project")
```


### Faux Desert High (Simulation of an Add Health HS)
```{r,fig.align='center',echo=FALSE,message=FALSE,out.width='.9\\textwidth'}
data(faux.desert.high)
plot(faux.desert.high,vertex.col="grade",edge.col=rgb(0,0,0,.3),displayisolates=FALSE)
col<-colorLegend(faux.desert.high,"grade")
legend("bottomright",col=col[order(col[,2]),2],pch=19,legend=col[order(col[,2]),1],bty="n")
```

\tiny{Hunter, D. R., Handcock, M. S., Butts, C. T., Goodreau, S. M., \& Morris, M. (2008). ergm: A package to fit, simulate and diagnose exponential-family models for networks. Journal of statistical software, 24(3).}


### World Trade Center Radio Communication
```{r,fig.align='center',echo=FALSE,message=FALSE,out.width='.5\\textwidth'}
load("data/relevent_sunbelt_2014.Rdata")
WTCPoliceNet <- as.sociomatrix.eventlist(WTCPoliceCalls,37) 
gplot(WTCPoliceNet,edge.lwd=WTCPoliceNet^0.75,vertex.col=2+WTCPoliceIsICR,vertex.cex=1.25,edge.col=rgb(0,0,0,.5))
title(main="Responder Communication Network")
```

\tiny{Butts, C. T., Petrescu-Prahova, M., \& Remy Cross, B. (2007). Responder communication networks in the World Trade Center disaster: Implications for modeling of communication within emergency settings. Mathematical Sociology, 31(2), 121-147.}

### David Krackhardt's Perceived Friendships

```{r,fig.align='center',echo=FALSE,message=FALSE,out.width='.7\\textwidth'}
data(krack)
kfr.conc<-consensus(krack$krackfr,method="romney.batchelder",verbose=FALSE)
##Plot
gplot(kfr.conc,displaylabels=TRUE,displayisolates = FALSE)
title("Consensus Model of Perceived Friendships in an Organization")
```

\tiny{Krackhardt, D. (1987). Cognitive social structures. Social networks, 9(2), 109-134.}

### SAR EMONs, All Reported Ties, from Drabek et al 1981
```{r,fig.align='center',echo=FALSE,message=FALSE,out.width='.8\\textwidth'}
data(emon)   #Load the emon data set
#Plot the EMONs
par(mfrow=c(2,4),mar=c(0, 0, .6, 0) + 0.1)
for(i in 1:length(emon)){
  plot(emon[[i]],edge.lwd="Frequency",edge.col=rgb(0,0,0,.4))
	title(main=names(emon)[i],outer=FALSE)
}
```

\tiny{Marcum, C. S., Bevc, C. A., \& Butts, C. T. (2012). Mechanisms of control in emergent interorganizational networks. Policy Studies Journal, 40(3), 516-546.}


### Needle Network

```{r,echo=FALSE,fig.align='center',results='hide',cache=TRUE}
load("data/HealthNetworks/flagstaff_rural.rda")
```

```{r,echo=FALSE,fig.align='center',out.width='.6\\textwidth'}
plot(flag_needle_net,vertex.col="race",main="Rural Arizona Risk Networks: Needle Sharing")
col<-colorLegend(flag_needle_net,"race")
legend("bottomright",col=col[,2],pch=19,legend=col[,1],bty="n")
```

\tiny{Morris, Martina, and Rothenberg, Richard. HIV Transmission Network Metastudy Project: An Archive of Data From Eight Network Studies, 1988--2001. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2011-08-09. https://doi.org/10.3886/ICPSR22140.v1}

### Sexual Contact Network

```{r,echo=FALSE,fig.align='center',out.width='.6\\textwidth'}
plot(flag_sex_net,vertex.col="race",main="Rural Arizona Risk Networks: Sex")
col<-colorLegend(flag_sex_net,"race")
legend("bottomright",col=col[,2],pch=19,legend=col[,1],bty="n")
```

\tiny{Morris, Martina, and Rothenberg, Richard. HIV Transmission Network Metastudy Project: An Archive of Data From Eight Network Studies, 1988--2001. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2011-08-09. https://doi.org/10.3886/ICPSR22140.v1}

### Social Contact Network

```{r,echo=FALSE,fig.align='center',out.width='.6\\textwidth'}
plot(flag_social_net,vertex.col="race",main="Rural Arizona Risk Networks: Social")
col<-colorLegend(flag_social_net,"race")
legend("bottomright",col=col[,2],pch=19,legend=col[,1],bty="n")
```

\tiny{Morris, Martina, and Rothenberg, Richard. HIV Transmission Network Metastudy Project: An Archive of Data From Eight Network Studies, 1988--2001. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2011-08-09. https://doi.org/10.3886/ICPSR22140.v1}

### Drug Contact Network

```{r,echo=FALSE,fig.align='center',out.width='.6\\textwidth'}
plot(flag_drug_net,vertex.col="race",main="Rural Arizona Risk Networks: Drug")
col<-colorLegend(flag_drug_net,"race")
legend("bottomright",col=col[,2],pch=19,legend=col[,1],bty="n")
```

\tiny{Morris, Martina, and Rothenberg, Richard. HIV Transmission Network Metastudy Project: An Archive of Data From Eight Network Studies, 1988--2001. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2011-08-09. https://doi.org/10.3886/ICPSR22140.v1}

# Network Representations (Data Structures/Visualizations)

### Defining Terms: A Basis in Graph Theory

- \red{Points} (or \red{nodes}/\red{vertices})
    - In social science – "actors"
        - People, organizations, nation-states, mobs, papers, ideas
    - In other settings – "entities"   
        - Animals, locations, objects

- \red{Lines} (or \red{ties}/\red{relations}) – interaction b/t nodes
    - These can be directed (\red{arcs}), non-directed (\red{edges}) and/or valued (counts/weights)
    - Social:
        - Friend, relative, neighbor, conversation partner, email sender/recipient, co-member, citations
    - Other:
        - Grooming behavior, traversal patterns, etc.

\tiny{Borgatti, SP. 1994. "A Quorum of Graph Theoretic Concepts." Connections 17:47-49.}

### Defining Terms: A Basis in Graph Theory

\begin{textblock*}{60mm}(-3mm,-.4\textheight)
\textbf{Network, graph, sociogram}
\begin{itemize}
\item Aggregate of ALL nodes AND ties for entire population in a single representation
\begin{itemize}
\item directed (digraph) or un-directed (graph)
\item Can be dissected into a variety of \red{subgraphs}
\end{itemize}
\item Mathematical notation, $G=(V,E)$
\end{itemize}
\end{textblock*}

\begin{textblock*}{55mm}(60mm,-.4\textheight)
\includegraphics[width=1\linewidth]{figures/graphSubgraph.png}
\end{textblock*}

\begin{textblock*}{100mm}(-3mm,.3\textheight)
\tiny{Borgatti, SP. 1994. "A Quorum of Graph Theoretic Concepts." Connections 17:47-49.}
\end{textblock*}

### Relational Data

\begin{textblock*}{50mm}(-3mm,-.3\textheight)
\tiny
```{r,echo=FALSE,results='asis'}
set.seed(1982)
g <-rgraph(5)
print(xtable(g,digits=0),floating=FALSE)
```
\end{textblock*}
\begin{textblock*}{40mm}(40mm,-.4\textheight)
```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='left'}
sociomatrixplot(g,cex.lab=3)
```
\end{textblock*}
\begin{textblock*}{30mm}(80mm,-.4\textheight)
```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='right'}
gplot(g)
```
\end{textblock*}

\begin{textblock*}{100mm}(-5mm,0\textheight)
\begin{itemize}
\item A collection of entities and a set of measured relations between them
\begin{itemize}
    \item \red{Entities}: nodes, actors, egos, units, respondents
    \item \red{Relations}: ties, links, edges
\end{itemize}
\item Relations can be
\begin{itemize}
    \item Directed or undirected
    \item Valued or dichotomous (binary)
\end{itemize}
\end{itemize}
\end{textblock*}

### Relational Data

\begin{textblock*}{50mm}(-3mm,-.3\textheight)
\tiny
\textbf{Adjacency Matrix}
```{r,echo=FALSE,results='asis'}
adj<-matrix(c(NA,0,1,0,0,0,0,NA,1,0,0,0,1,1,NA,1,0,0,0,0,1,NA,1,1,0,0,0,1,NA,1,0,0,0,1,1,NA),nc=6,nr=6)
rownames(adj)<-colnames(adj)<-1:6
print(xtable(adj,digits=0),floating=FALSE)
```

\begin{itemize}
\item Self-ties are known as "loops"
\item Typically ignored
\end{itemize}

\end{textblock*}
\begin{textblock*}{40mm}(40mm,-.4\textheight)
\tiny
\textbf{Edge List (Arc List)}

```{r,echo=FALSE,results='asis'}
el<-as.edgelist.sna(adj,suppress.diag =TRUE)
print(xtable(el[,1:2],digits=0),floating=FALSE,include.rownames=FALSE)
```
\end{textblock*}
\begin{textblock*}{50mm}(60mm,-.4\textheight)
\tiny
\begin{center}
\textbf{Graph}
\end{center}
```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='right'}
set.seed(99)
gplot(adj,label=colnames(adj),gmode="graph",vertex.cex=1.5,label.cex=2)
```
\end{textblock*}

\begin{textblock*}{100mm}(-5mm,0.4\textheight)
\tiny{Knoke D, Yang S. Social Network Analysis: Sage; 2007.}
\end{textblock*}

### Relational Data

\begin{textblock*}{50mm}(-3mm,-.3\textheight)
\tiny
\textbf{Friendship Nomination}

\textbf{Adjacency Matrix}

```{r,echo=FALSE,results='asis'}
adj<-matrix(c(NA,2,0,0,0,0,1,NA,1,0,0,0,1,1,NA,2,0,0,0,0,2,NA,0,1,0,0,0,1,NA,2,0,0,0,0,0,NA),nc=6,nr=6)
rownames(adj)<-colnames(adj)<-1:6
print(xtable(adj,digits=0),floating=FALSE)
```

\end{textblock*}
\begin{textblock*}{40mm}(40mm,-.4\textheight)
\tiny
\textbf{Friendship Nomination}

\textbf{Edge List (Arc List)}

```{r,echo=FALSE,results='asis'}
el<-as.edgelist.sna(adj,suppress.diag =TRUE)
print(xtable(el[!is.na(el[,3]),1:2],digits=0),floating=FALSE,include.rownames=FALSE)
```
\end{textblock*}
\begin{textblock*}{50mm}(60mm,-.4\textheight)

\begin{center}
\tiny
\textbf{Friendship Nomination}

\textbf{Graph}
\end{center}

```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='right'}
set.seed(9)
gplot(adj,label=colnames(adj),gmode="digraph",vertex.cex=1.5,label.cex=2,usecurve =TRUE)
```
\end{textblock*}

\begin{textblock*}{100mm}(-5mm,0.4\textheight)
\tiny{Knoke D, Yang S. Social Network Analysis: Sage; 2007.}
\end{textblock*}

### Relational Data

\begin{textblock*}{50mm}(-3mm,-.3\textheight)
\tiny
\textbf{Friendship Nomination}

\textbf{Adjacency Matrix (Weighted)}

```{r,echo=FALSE,results='asis'}
adj<-matrix(c(NA,2,0,0,0,0,1,NA,1,0,0,0,1,1,NA,2,0,0,0,0,2,NA,0,1,0,0,0,1,NA,2,0,0,0,0,0,NA),nc=6,nr=6)
rownames(adj)<-colnames(adj)<-1:6
print(xtable(adj,digits=0),floating=FALSE)
```

\end{textblock*}
\begin{textblock*}{40mm}(40mm,-.4\textheight)
\tiny
\textbf{Friendship Nomination}

\textbf{Weighted Edge List (Arc List)}

```{r,echo=FALSE,results='asis'}
el<-as.edgelist.sna(adj,suppress.diag =TRUE)
print(xtable(el[!is.na(el[,3]),1:3],digits=0),floating=FALSE,include.rownames=FALSE)
```
\end{textblock*}
\begin{textblock*}{50mm}(60mm,-.4\textheight)
\begin{center}
\tiny
\textbf{Friendship Nomination}

\textbf{Graph}
\end{center}

```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='right'}
set.seed(9)
gplot(adj,edge.lwd=5*adj,label=colnames(adj),gmode="digraph",vertex.cex=1.5,label.cex=2,usecurve =TRUE)
```
\end{textblock*}

\begin{textblock*}{100mm}(-5mm,0.4\textheight)
\tiny{Knoke D, Yang S. Social Network Analysis: Sage; 2007.}
\end{textblock*}

### Relational Data

\begin{textblock*}{50mm}(-3mm,-.3\textheight)
\tiny
\textbf{Friendship Nomination}

\textbf{Adjacency Matrix (Weighted)}
```{r,echo=FALSE,results='asis'}
adj<-matrix(c(NA,2,0,0,0,0,1,NA,1,0,0,0,1,1,NA,2,0,0,0,0,2,NA,0,1,0,0,0,1,NA,2,0,0,0,0,0,NA),nc=6,nr=6)
rownames(adj)<-colnames(adj)<-1:6
print(xtable(adj,digits=0),floating=FALSE)
```

\tiny
\textbf{\red{Family Nomination}}
```{r,echo=FALSE,results='asis'}
adjf<-matrix(c(NA,1,1,0,0,0,1,NA,1,0,0,0,1,1,NA,0,0,0,0,0,0,NA,1,1,0,0,0,1,NA,1,0,0,0,1,1,NA),nc=6,nr=6)
rownames(adjf)<-colnames(adjf)<-1:6
adjfp<-matrix(paste("\\red{",adjf,"}",sep=""),nc=6,nr=6)
rownames(adjfp)<-colnames(adjfp)<-paste("\\red{",1:6,"}",sep="")
diag(adjfp)<-""
print(xtable(adjfp,digits=0),floating=FALSE,sanitize.text.function = function(x) x)
```

\end{textblock*}
\begin{textblock*}{40mm}(40mm,-.4\textheight)
\end{textblock*}

\begin{textblock*}{60mm}(50mm,-.4\textheight)
\begin{center}
\tiny
\textbf{Friendship Nomination}
\end{center}

```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='right'}
set.seed(9)
coords<-gplot(adj,edge.lwd=5*adj,label=colnames(adj),gmode="digraph",vertex.cex=1.5,label.cex=2,usecurve =TRUE)
gplot(adjf,new=FALSE,coord=coords,vertex.cex=1.5,edge.col="red")
```
\end{textblock*}

\begin{textblock*}{100mm}(-5mm,0.4\textheight)
\tiny{Knoke D, Yang S. Social Network Analysis: Sage; 2007.}
\end{textblock*}

### Relational Data: Multiplexity

\begin{center}
\includegraphics[width=.85\linewidth]{figures/multiplexTies}
\end{center}

\tiny{adams j, Moody J, Morris M. "Sex, Drugs, and Race:  How Behaviors Differentially Contribute to Sexually Transmitted Infection Risk Network Structure." AJPH 2013;103(2):322-9.}

### Relational Data: Network Change

\begin{center}
\includegraphics[width=.7\linewidth]{figures/dynamics.png}
\end{center}

\tiny{Butts, C. T., Acton, R. M., \& Marcum, C. S. (2012). Interorganizational Collaboration in the Hurricane Katrina Response. Journal of Social Structure, 13.}

### Common Features of Interest

Beyond nodal and dyadic attributes, many networks exhibit the following features:

- \red{Reciprocity} of ties
- \red{Degree heterogeneity} among actors 
    - Sociability, Popularity
- \red{Homophily} by actor attributes
    - Higher propensity to form ties between actors with similar attr
- \red{Transitivity} of relationships
    - Friends of friends have a higher propensity to be friends
- \red{Balance} of relationships
    - Liking those who dislike whom you dislike
- \red{Equivalence} of nodes
    - Some nodes may have identical/similar patterns of relationships

### How to Think about Network Data

\begin{center}
\includegraphics[width=1\linewidth]{figures/typology}
\end{center}

\tiny{James Moody."Introduction to Social Network Analysis - Duke Sociology"}

### How to Think about Network Data

\begin{center}
\includegraphics[width=.8\linewidth]{figures/networkdata}
\end{center}

### Frameworks (Orientation)

- Methodological Aims
   - Level of Analysis
   - Unit of Analysis
- Theoretical Modtivations
    - Descriptive
    - Analytic


\tiny{Knoke D, Yang S. Social Network Analysis: Sage; 2007.}


### Level of Analysis: Sampling

- Ego-Networks
    - A respondent and the set of people they have relationships with
- Measures:
    - Size
    - Density
    - Types of Relationships
    - Similarity
    - Composition

\begin{textblock*}{60mm}(50mm,-.4\textheight)
```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='right',cache=TRUE}
load("data/egonet.Rda")

build_edgelist_sna<-function(el){
vnam<-unique(unlist(el))
vnam<-vnam[order(vnam)]
e1<-match(el[,1],vnam)
e2<-match(el[,2],vnam)
el_sna<-cbind(e1,e2,rep(1,length(e1)))
colnames(el_sna)<-c("snd","rec","weight")
attr(el_sna,"n")<-length(vnam)
attr(el_sna,"vnames")<-vnam
el_sna
}

el_sna<-build_edgelist_sna(el=egonet[,1:2])
mat<-as.sociomatrix.sna(el_sna)
mat[,]<-symmetrize(as.sociomatrix.sna(el_sna))
col<-rep("red",nrow(mat))
col[rownames(mat)=="Me"]<-"blue"
par(mar=c(0, 0, 0, 0) + 0.1)
gplot(mat,gmode="graph",vertex.col=col,label=rownames(mat),label.cex=1.5)
```
\end{textblock*}

\begin{textblock*}{60mm}(10mm,.5\textheight)
\tiny{Laumann EO, Marsden PV, Prensky D. 1994. "The Boundary Specification Problem in Network Analysis." In: Freeman LC, White DR, Romney AK, (eds.)}
\end{textblock*}

### Levels of Analysis: Complete Population

- Complete or Global Networks
    - The collection or relationships among all members of a population
- Measures:
    - Ego Network (collection of all egos/respondents)
    - Centrality
    - Sub-groups (sometimes called clusters or communities)
    - Equivalence (and orther issues of role)
    
### Levels of Analysis: Why do levels matter?

**Who’s at greatest risk of contracting an STI?**
\begin{center}
\includegraphics[width=.6\linewidth]{figures/stiexample}
\end{center}

\begin{textblock*}{80mm}(10mm,0\textheight)
\tiny{Bearman PS, Moody J, Stovel K. Chains of affection: The structure of adolescent romantic and sexual networks. American Journal of Sociology 2004;110(1):44-9}
\end{textblock*}

### Levels of Analysis: Why do levels matter?

**Who’s at greatest risk of contracting an STI?**

\begin{center}
\includegraphics[width=.6\linewidth]{figures/stiexample2}
\end{center}

\begin{textblock*}{80mm}(10mm,0\textheight)
\tiny{Bearman PS, Moody J, Stovel K. Chains of affection: The structure of adolescent romantic and sexual networks. American Journal of Sociology 2004;110(1):44-9}
\end{textblock*}

### Unit of Analysis: Measurement Issues

\begin{textblock*}{80mm}(0mm,-.4\textheight)
\begin{itemize}
\item Measurement issues can occure at varying levels
\begin{itemize}
\item Nodes/Vertices/Actors
\item Ties/Relationships/Edges
\item (small) Groups
\item Complete Networks
\end{itemize}
\end{itemize}

\begin{center}
\textbf{Who/What is Important?}
\end{center}
\end{textblock*}

\begin{textblock*}{60mm}(60mm,-.1\textheight)
```{r,echo=FALSE,out.width='.9\\textwidth',fig.align='left',cache=TRUE}
data(emon)
i<-6
plot(emon[[i]],
sub=names(emon)[i],
edge.lwd="Frequency",
displayisolates=FALSE, 
vertex.col=rainbow(8),displaylabels=FALSE,
edge.col=rgb(0,0,0,.3))
legend("bottomright",legend=(emon[[i]]%v%"vertex.names")[sna::degree(emon[[i]])!=0],pch=19,col=rainbow(8),bty="n",cex=.7)
```
\end{textblock*}

### Theoretical Frameworks: Common Metaphors

**"Pipes" – A Connectionist Metaphor**

- Networks provide potential pathways through which various ``bits"" can be passed from one node to another via the links between them, e.g.:
    - Diffusion of information
    - Spread of an infectious disease
    - Traffic flows
    - VoIP
- Analyzing networks therefore aims to identify properties that will \red{promote} or \red{constrain} the potential transmission of those bits through the population
    - Node-level (trust)
    - Dyad-level (respect, distance, status difference, etc.)
    - Network-level (centrality, clustering, timing, etc.)

\tiny{adams, jimi. EPIC - SNA, 2017. Columbia University.}

### Theoretical Frameworks: Common Metaphors

**"Pipes" – A Connectionist Metaphor**

  - [A fun game!](http://vax.herokuapp.com/) to Illustrate the ideas

\tiny{adams, jimi. EPIC - SNA, 2017. Columbia University.}

### Theoretical Frameworks: Common Metaphors

**"Prisms" - A Topological Metaphor**

- Network patterns reveal \red{differences} or \red{similarities} in \red{status/role} comparisons
    - "equivalence" (e.g., relatives, employment hierarchies)
    - Identify groups
    - Identify advantageous individual positions
    - Role/identity construction – e.g., through shared memberships

- SNA in this case is interested in identifying patterns that can \red{differentiate} between varieties of \red{networks} or varieties of \red{locations} within networks
    - Equivalence 
    - Cohesion
    - Clustering
    
\tiny{adams, jimi. EPIC - SNA, 2017. Columbia University.}

### Theoretical Frameworks: Common Metaphors

- Why might a pair of nodes be "similar"? 
    - Pipes metaphor
        - Connectionist/Pathways 
    - Prisms metaphor
        - Topological/Roles

\tiny{adams, jimi. EPIC - SNA, 2017. Columbia University.}

### Theoretical Frameworks: Common Metaphors

\includegraphics[width=.8\linewidth]{figures/metaphor}

\begin{center}
\tiny{adams, jimi. EPIC - SNA, 2017. Columbia University.}
\end{center}


### Common Features of Interest

Beyond nodal and dyadic attributes, many networks exhibit the following features:

- \red{Reciprocity} of ties
- \red{Degree heterogeneity} among actors 
    - Sociability, Popularity
- \red{Homophily} by actor attributes
    - Higher propensity to form ties between actors with similar attr
- \red{Transitivity} of relationships
    - Friends of friends have a higher propensity to be friends
- \red{Balance} of relationships
    - Liking those who dislike whom you dislike
- \red{Equivalence} of nodes
    - Some nodes may have identical/similar patterns of relationships

\tiny{Peter Hoff. Statistical Networks. Lecture Notes. University of Washington.}

# Review of R and Discussion of Network Data in R

### Review of R

- [Intro R Tutorial](https://ssdalab.github.io/IntroR/)


# References and Places for More Information 

### References and Places for More Information {.allowframebreaks}


