---
title: "Social Network Analysis: Descriptives Part 2"
subtitle: "EPIC - SNA, Columbia University"
author: "Zack W Almquist"
institute: "University of Minnesota"
date: "June 12th, 2018"
output: 
  beamer_presentation:
    includes:
      in_header: header.tex
      after_body: footer.tex
toc: true
#toc_depth: 2
#keep_tex: true
#https://github.com/matze/mtheme
---

```{r,include=FALSE,cache=FALSE}
require(knitr)
opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE,tidy=TRUE,comment=NA,prompt=FALSE)
options(width=40)

## Packages
library(networkMethods)
library(networkdata)
library(sna)
library(network)
library(ergm)
library(xtable)
```


# Degree Distribution

### Degree Distribution: Local Level

Given some ego, what do we want to know about their alters?

- Network Size
    - Degree - total number of ties incident to a node
        - E.g., GSS "Important matters" networks: \red{2.9}(1985);\red{2}(2004)
        - E.g, "Dunbar Number": 150-200

```{r,echo=FALSE,message=FALSE,fig.align='center',out.width='.3\\textwidth',cache=TRUE}
betex.star<-matrix(0,nc=6,nr=6)
betex.star[1,2:6]<-1
betex.star[2:6,1]<-1
gplot(betex.star,gmode="graph",vertex.cex = 2)
```    

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}


### Degree Estimation

Not just central tendencies, but distribution(s) too.   

- How many friends do most people have?

vs.

- How many friends "should" I have?

```{r,echo=FALSE,fig.align='center',out.width='.4\\textwidth'}
hist(rpois(10000,5))
```

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}

### Degree Estimation

\begin{center}
\includegraphics[width=1\linewidth]{figures/feld}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Feld SL. Why Your Friends have More Friends than you do. American Journal of Sociology 1991;96:1464-1477.}

### Degree Estimation

\begin{center}
\includegraphics[width=1\linewidth]{figures/feld2}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Feld SL. Why Your Friends have More Friends than you do. American Journal of Sociology 1991;96:1464-1477.}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg17}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg18}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg16}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Barabasi AL, Albert R, Jeong H, Bianconi G. Power-Law Distribution of the World Wide Web. Science 2000;287:2115.}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg15}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Barabasi AL, Albert R, Jeong H, Bianconi G. Power-Law Distribution of the World Wide Web. Science 2000;287:2115.}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg14}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:Butler RJ, Cowan BW, Nilsson S. From Obscurity to Bestseller: Examining the Impact of Oprah's Book Club Selections. Publishing Research Quarterly 2005;20(4):23-34}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg13}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Dezso Z, Barabasi AL. Halting viruses in scale-free Networks. Physical Review E 2002.}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg12}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Dezso Z, Barabasi AL. Halting viruses in scale-free Networks. Physical Review E 2002.}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg11}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Amaral LAN, Scala A, Barthelemy M, Stanley HE. Classes of Small World Networks. PNAS 2000;97(21):1}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg10}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:}

### Degree Distributions Matter

**Adults & Children Estimated to be Living w/HIV, 2011**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg9}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg8}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:May RM, Anderson RM. Transmission Dynamics of HIV Infection. Nature 1987;326:137-142.}

### Degree Distributions Matter

**Who is at greatest/least risk?**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg7}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:}

### Degree Distributions Matter

**STI Interventions**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg6}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg5}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Liljeros F, Edling CR, Amaral LAN, Stanley HE, Aberg Y. The Web of Human Sexual Contacts. Nature 2001;411:9}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg4}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Morris (2003)}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg3}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Morris (2003)}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg2}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Galvin SR, COhen MS. The Role of Sexually Transmitted Diseases in HIV Transmission. Nature Reviews: Microbiology 2004;2:33-42.}

### Degree Distributions Matter

\begin{center}
\includegraphics[width=.8\linewidth]{figures/deg1}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin:}

# Network Composition

### Homophily: Add Health Friendship Network

\begin{textblock*}{100mm}(10mm,-.4\textheight)
```{r,warning=FALSE,message=FALSE,out.width='1\\linewidth',echo=FALSE}
library(networkdata)
data(addhealth)
plot(addhealth[[10]],vertex.col="grade",displayisolates=FALSE,edge.col=rgb(0,0,0,.1),vertex.cex=.5)
legend("bottomleft",legend=unique(addhealth[[10]]%v%"grade")[order(unique(addhealth[[10]]%v%"grade"))],col=unique(addhealth[[10]]%v%"grade")[order(unique(addhealth[[10]]%v%"grade"))],pch=19,title="Grade")
```
\end{textblock*}


### Selective Mixing

\begin{textblock*}{70mm}(0mm,-.4\textheight)
\begin{itemize}
\item Observation: vertices in most real-world networks are heterogeneous
\begin{itemize}
\item Different classes of vertices unequally likely to be tied
\end{itemize}
\item Selective mixing: differences in marginal tie probability among groups
\end{itemize}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}

\begin{textblock*}{55mm}(60mm,-.2\textheight)
```{r,warning=FALSE,message=FALSE,out.width='1\\linewidth',echo=FALSE}
i<-1
plot(addhealth[[i]],vertex.col="sex",displayisolates=FALSE,edge.col=rgb(0,0,0,.1),vertex.cex=.8)
legend("bottomleft",legend=c("Male","Female"),col=unique(addhealth[[i]]%v%"sex")[order(unique(addhealth[[i]]%v%"sex"))],pch=19,title="Gender")
```
\end{textblock*}


### Homophily

\begin{textblock*}{70mm}(0mm,-.4\textheight)
\begin{itemize}
\item Very common type of selective mixing: marginal tendency of individuals with similar properties to be adjacent
\begin{itemize}
\item Term refers to association, not underlying mechanism
\end{itemize}
\item Can occur for many reasons
\begin{itemize}
\item Active selection (i.e., preferences)
\item Passive selection (e.g., tracking in schools)
\item Diffusion/influence
\end{itemize}
\item Less well-understood than one would think
\end{itemize}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}

\begin{textblock*}{55mm}(60mm,-.2\textheight)
```{r,warning=FALSE,message=FALSE,out.width='.9\\linewidth',echo=FALSE}
plot(addhealth[[1]],vertex.col="grade",displayisolates=FALSE,edge.col=rgb(0,0,0,.1),vertex.cex=.8)
legend("bottomleft",legend=unique(addhealth[[1]]%v%"grade")[order(unique(addhealth[[1]]%v%"grade"))],col=unique(addhealth[[1]]%v%"grade")[order(unique(addhealth[[1]]%v%"grade"))],pch=19,title="Grade")
```
\end{textblock*}



### Propinquity and Age Mixing

\begin{textblock*}{70mm}(0mm,-.4\textheight)
\begin{itemize}
\item More complex mixing is also possible
\item Propinquity: tendency of spatially proximate vertices to be tied
\begin{itemize}
\item Continuous covariate -- function form of effect important for resulting structure
\end{itemize}
\item Age mixing: differential association based on age 
\begin{itemize}
\item  Often broadly homophilous, but details can vary
\item Asymmetries common, e.g., male/female age difference in marriages
\end{itemize}
\end{itemize}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}
\end{textblock*}

\begin{textblock*}{55mm}(60mm,0\textheight)
```{r,warning=FALSE,message=FALSE,out.width='1\\linewidth',echo=FALSE}
load("data/beach.rda")
par(mfrow=c(1,1),mar=c(0, 0, 0, 0) + 0.3)
for(i in 4){
	if(is.network(beach[[i]])){
gplot(beach[[i]],coord=beach[[i]]%n%"cord",vertex.cex=.4,usearrows=FALSE,edge.lwd=.5,edge.col=rgb(0,0,0,0.1),vertex.col=beach[[i]]%v%"regular"+1)
mtext(paste(substr(names(beach)[i],1,1),"/",substr(names(beach)[i],2,3),"/1987",sep=""),side=1,cex=.5)
}
}
```
\end{textblock*}


### Freeman et al. (1988)

```{r,warning=FALSE,message=FALSE,out.width='1\\linewidth',echo=FALSE}
load("data/beach.rda")
par(mfrow=c(3,10),mar=c(0, 0, 0, 0) + 0.3)
for(i in 1:length(beach)){
	if(is.network(beach[[i]])){
gplot(beach[[i]],coord=beach[[i]]%n%"cord",vertex.cex=.4,usearrows=FALSE,edge.lwd=.5,edge.col=rgb(0,0,0,0.1),vertex.col=beach[[i]]%v%"regular"+1)
mtext(paste(substr(names(beach)[i],1,1),"/",substr(names(beach)[i],2,3),"/1987",sep=""),side=1,cex=.5)
}
}
```


### Mixing Rates: Some Basics

\begin{itemize}
\item Observation: we already know how to calculate one kind of mixing rate -- the density!
\begin{itemize}
\item Rate at which all vertices mix with all other vertices
\end{itemize}
\item Generalization to multiple groups
\begin{itemize}
\item Let $S$, $S^{'}$ be two groups. Then
$$r_{SS^{'}} = \frac{\sum_{i\in S}\sum_{j\in S^{'}}Y_{ij}}{|\{ (s,s^{'}):s\in S,s^{'}\in S^{'}, s\neq s^{'})\}|}$$
\item Not that this reduces to density when $S=S^{'}=V$ (loops omitted here)
\end{itemize}
\item Relationship to inhomogeneous Bernoulli Model
\begin{itemize}
\item For $i\in S$, $j\in S^{'}$ let $\Phi_{ij}=r_{SS^{'}}$. Bernoulli model with parameter $\Phi$ expresses distribution in which edges are independent and homogeneous within groups, with fixed inter/intra-group mixing rates 
\end{itemize}
\end{itemize}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}

### Null Distributions
\begin{itemize}
\item Several ways to obtain null distributions
\begin{itemize}
\item Binomial model
\begin{itemize}
\item Let $X_{SS^{'}}$ be the number of edges from $S$ to $S^{'}$, and let $X_{SS^{'}}^{*}$ be the number of possible edges. If all $S,S^{'}$ edge variables are independent with probability $p$, then $X_{SS^{'}} \sim Binom(X_{SS^{'}}^{*} ,p)$; its expected value is thus $pX_{SS^{'}}^{*}$ and its variance is $p(1-p)X_{SS^{'}}^{*}$
\item Can get null from assumption that $p$ is equal to the density
\end{itemize}
\item Poisson approximation
\begin{itemize}
\item If $X_{SS^{'}}^{*} \rightarrow \infty$ with $pX_{SS^{'}}^{*}$ fixed, then $X_{SS^{'}}^{*}\sim Pois(pX_{SS^{'}}^{*})$; expected value and variance in this case are both $pX_{SS^{'}}^{*}$
\item Can get null expectation using two-way marginals: $pX_{SS^{'}}^{*}=X_S X_{S^{'}}/(\sum_i\sum_j Y_{ij})$; comparison using z-scores typically works fairly well for exploratory purposes
\end{itemize}
\item Permutation of vertex labels
\begin{itemize}
\item Randomly re-assign set memberships, holding set sizes constant, and recalculate $X_{SS^{'}}$. This both exact, and preserves edge value distributions (useful if valued)
\end{itemize}
\end{itemize}
\end{itemize}

\tiny{Carter Butts. Social Network Methods. University of California, Irvine.}

# Network Position (Theories)

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh11}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Burt RS. “Structural Holes versus Network Closure as Social Capital.” Pp.31-56ininLinN,CookK,BurtRS.Social}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh10}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Burt RS. Structural Holes: The social structure of competition. Harvard University Press; 1992.}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh9}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Rodan S. Structural Holes and Managerial Performance:Identifying Underlying  Mechanisms. Social Networks.2010;32(3).}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh8}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Burt RS. Secondhand Brokerage: Evidence on the Importance of Local Structure for Managers, Bankers, and Analysts.
Academy of Management Journal 2007;50(1):119-148.}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh7}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh6}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh5}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Burt RS. Structural Holes: The social structu}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh4}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Burt RS. Structural Holes: The social structure of competition. Harvard University Press; 199}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh3}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Granovetter M. The Strength of Weak Ties. American Journal of Sociology 1973;81:1287-1303.}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh2}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Jim Moody}

### Structural Holes

\begin{center}
\includegraphics[width=.8\linewidth]{figures/sh1}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

# Distance Measures

### Distance Measures (reachability)

\begin{center}
\includegraphics[width=1\linewidth]{figures/dist21}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Milgram S. The Small World Problem. Psychology Today 1967;1:61-67}

### Distance Measures (reachability)

**The Small World of Facebook**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist20}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{[Facebook](https://www.facebook.com/notes/facebook-data-team/anatomy-of-facebook/10150388519243859)}

### Distance Measures (reachability)

\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist19}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist18}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist17}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist16}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist15}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist14}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist13}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist12}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist11}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Geodesics**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist10}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Distance Matters**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist9}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Distance Matters**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist8}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Incomplete Path Lengths**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist7}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Is 6 large or small number?**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist6}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Direct Searches**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist5}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{}

### Distance Measures (reachability)

**Why were they so far off?**
\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist4}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Killworth PD, McCarty C, Bernard HR, House M. The Accuracy of Small World Chains in Social Networks. Social Networks 2006;28.}

### Distance Measures (reachability)


\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist3}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Milgram S. The Small World Problem. Psychology Today 1967;1:61-6}.}

### Distance Measures (reachability)

\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist2}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Milgram S. The Small World Problem. Psychology Today 1967;1:61-6}

### Distance Measures (reachability)

\begin{center}
\includegraphics[width=.8\linewidth]{figures/dist1}
\end{center}

\tiny{jimi adams. EPIC- SNA 2017. Columbia University.}
\tiny{Origin: Milgram S. The Small World Problem. Psychology Today 1967;1:61-6}


# R for Descriptive Analysis: More tools

### References and Places for More Information {.allowframebreaks}

